---
title: "Personalize your own ggplot theme"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(2022) # so we all get the same simulated data set
```

Please use this file to code along during the workshop.
Speak up as soon as possible if something does not work for you!
We are happy to help you out.

## create data set
First, we create a data set to illustrate our example. 
We simulate 100 patients who were included into the registries between 2000 and now.

```{r libraries, warning=FALSE, message=FALSE, echo = FALSE}
# if you have not yet installed all of these libraries:
# run install.packages("ggplot2") etc for each of the packages you need

library(dplyr)
library(tidyr) # complete()
library(lubridate) # floor_date()
library(ggplot2)
```

```{r data}
# we simulate a registry which was started in 2000 and by now has 100 patients
scqm_data <- data.frame(
  disease = sample(factor(c("RA", "axSpA","PsA"), 
                          levels = c("RA", "axSpA", "PsA")), 
                   size = 100, 
                   replace = TRUE, 
                   prob = c(0.5, 0.3, 0.2)),
  gender = factor(sample(x = c("m", "f"), 
                         size = 100, replace = TRUE), 
                  levels = c("m", "f")),
  inclusion_date = as.Date(seq(from = as.Date("2000-01-01"), 
                               to = Sys.Date(), 
                               length.out = 100))
)
summary(scqm_data)

# the summarized data set for the plot: 
# number of patients per year, gender and disease
dat_sum <- scqm_data %>% 
  mutate(inclusion_year = floor_date(ymd(inclusion_date), unit = "year")) %>%
  group_by(inclusion_year, disease, gender) %>%
  summarise(n = n(), .groups = "drop") %>%
  complete(inclusion_year, disease, gender,  fill=list(n=0))  %>%
  arrange(disease, inclusion_year, gender) %>%
  group_by(disease, gender) %>%
  mutate(n_cumulative = cumsum(n)) %>%
  as.data.frame()
str(dat_sum)

# create labels for latest cumulative value
dat_sum <- dat_sum %>% mutate(label = case_when(
  (inclusion_year == max(dat_sum$inclusion_year)) ~ n_cumulative,
  TRUE ~ NA_integer_
))
```


## ggplot themes
### ggplot default theme
```{r original}
p <- ggplot(dat_sum, aes(x = inclusion_year, y = n_cumulative, fill = disease)) +
  geom_area(stat = "identity") +
  facet_wrap(~gender) +
  xlab("Year of inclusion") +
  ylab("Cumulative nr. of patients")
p
```


### other ggplot themes
Most of you probably know some of the ggplot built-in themes such as theme_bw() 
or theme_minimal(). The background changed from grey to white.
```{r mininal}
p + theme_bw()
```

### BBC theme
We try how our plot looks like with a theme that BBC shares
online: https://github.com/bbc/bbplot
We see some changes to the grid lines, and the legend is now on top.
```{r bbcstyle}
source("bbc_style.R")
p + bbc_style()
```

### SCQM theme
Now we want to create our own theme.
```{r scqm}
source("theme_scqm.R")
p + theme_scqm()

# an adapted version to show how you can still tweak it
p +
  theme_scqm(text_color = "#7b4264", grid_color = "#7b4264", 
             alignment = 0, font = "serif") +
  theme(legend.position = "top", legend.title = element_blank())
```

## custom color palettes
With the theme, we were able to manipulate most aspects of the plot, but not
the colors used in the plot. To define custom colors, I followed this post: https://meghan.rbind.io/blog/2022-10-11-creating-custom-color-palettes-with-ggplot2/

Take a look at the slides from R-ladies London on the topic of accessibility:
https://nrennie.rbind.io/talks/2022-october-rladies-london/


### defining the color palettes
We will create 2 functions: one which defines the HEX code of the colors we use
in our palettes. The second function defines the palettes we will want to use
later on in the plots.
```{r scqm-colors}
# show the palettes we created in theme_scqm.R
scales::show_col(scqm_palette("main"), cex_label = 2)
scales::show_col(scqm_palette("mainlight"), cex_label = 2)
scales::show_col(scqm_palette("repreg"), cex_label = 2)
scales::show_col(scqm_palette("repreglight"), cex_label = 2)
```


### defining ggplot color scales
We will now define functions which define custom ggplot color scales.
```{r scqm-scale-fill}
# with main palette
p +
  theme_scqm() +
  scale_fill_scqm()


# with repreg palette
p +
  theme_scqm() +
  scale_fill_scqm("repreg")


# with repreg light palette
p +
  theme_scqm() +
  scale_fill_scqm("repreglight")
```

If you do not want to use the colors of a palette but manually select your
choice of CICD colors, you can use scale_fill_manual().
```{r scale-fill-manual}
# with manually selected colors
p +
  theme_scqm()+
  scale_fill_manual(values =  
                      unname(c(scqm_color("grey", "lightsand", "lightrose"))))
```

### taking it further
Often, I want to keep the color for i.e. the different diseases 
consistent across my entire report. The first obvious choice is to use
a custom palette for diagnose.
```{r color-per-diagnose}
# palette 'diagnose'
p +
  theme_scqm() +
  scale_fill_scqm("diagnose")


# Drawback: color is assigned by level ordering (1st is RA 'blue', not RA)
# this only works if you order 'disease' in the same order as in the palette
# an example of where it does not work:
dat_sum_releveled <- dat_sum %>% mutate(disease = relevel(disease, ref = "PsA"))
ggplot(dat_sum_releveled, 
       aes(x = inclusion_year, y = n_cumulative, fill = disease)) +
  geom_area(stat = "identity") +
  facet_wrap(~gender)+
  theme_scqm() +
  scale_fill_scqm("diagnose")
```

I therefore played around with hardcoding the colors with a named vector.
This works nicely, if the vector has exactly the length and order of your levels.
```{r diagnose-solution}
# my best solution so far:
# even though RA is no longer the reference level, RA is still blue
diagnose_colors <- scqm_palette("diagnose")[levels(dat_sum_releveled$disease)]
ggplot(dat_sum_releveled, 
       aes(x = inclusion_year, y = n_cumulative, fill = disease)) +
  geom_area(stat = "identity") +
  facet_wrap(~gender)+
  theme_scqm() +
scale_fill_manual(values = diagnose_colors)


## a different color for gender
gender_colors <- scqm_color("mint", "sand")
names(gender_colors) <- c("m", "f")

ggplot(dat_sum, 
       aes(x = inclusion_year, y = n_cumulative, fill = gender)) +
  geom_area(stat = "identity") +
  facet_wrap(~disease)+
  theme_scqm() +
scale_fill_manual(values = gender_colors)
```

### a quick preview of other color scales
We now only showed examples of the color scale 'fill'.
Here are a few examples of other types of graphs.
```{r other-plot-types}
### line plot .............
ggplot(dat_sum, 
       aes(x = inclusion_year, y = n_cumulative, color = disease, label = label)) +
  geom_line() +
  # we don't want to show a warning for the missing values and we do not want 'a' in the legend
  geom_label(na.rm = TRUE, show.legend = FALSE) + 
  
  facet_wrap(~gender)+
  theme_scqm() +
  scale_color_scqm()


### dot plot .............
ggplot(dat_sum, 
       aes(x = inclusion_year, y = n_cumulative, color = disease, label = label)) +
  geom_point() +
  theme_scqm() +
  scale_color_scqm()


### box plot .............
ggplot(dat_sum, 
       aes(x = disease, y = n, fill = gender, label = label)) +
  geom_boxplot() +
  theme_scqm() +
  scale_fill_manual(values = gender_colors)
```



## global ggplot options
It is also possible to set the colors as default at the beginning of your report.
```{r}
# from https://stackoverflow.com/questions/10504724/change-the-default-colour-palette-in-ggplot
options(ggplot2.discrete.colour= unname(scqm_palette("main")))
options(ggplot2.discrete.fill= unname(scqm_palette("main")))

p
```

